---
title: "geneClusterPattern Vignette: overview"
author: "Jianhong Ou"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('geneClusterPattern')`"
abstract: >
  Visualize conserved gene clusters in multiple genomes
vignette: >
  %\VignetteIndexEntry{geneClusterPattern Vignette: overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    theme: simplex
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: true
---

```{r, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
    library(org.Dr.eg.db)
    library(geneClusterPattern)
    library(GenomeInfoDb)
    fish <- readRDS(system.file('extdata', 'fish.rds',
                            package = 'geneClusterPattern'))
    homologs <- readRDS(system.file('extdata', 'homologs.rds',
                                package = 'geneClusterPattern'))
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

# Introduction


Understanding the conservation of gene neighborhoods across species provides important insights into genome organization, functional linkage, and evolutionary processes. **geneClusterPattern** is a package developed to facilitate the systematic characterization of orthologous gene arrangements surrounding a focal gene across multiple species.

Given a gene of interest, the package supports two complementary strategies to define orthology relationships. It can retrieve homologous genes directly from Ensembl (via Compara), or it can incorporate user-provided ortholog group assignments generated by [OrthoFinder](https://github.com/davidemms/OrthoFinder). These flexible input options allow users to integrate gene homology information from widely used comparative genomics pipelines.

Based on the orthology information, **geneClusterPattern** constructs an ordered representation of genes within a configurable genomic window upstream and downstream of the focal gene. The resulting ortholog gene pattern captures the relative arrangement of ortholog groups across species, enabling direct comparison of local genomic structure.

This framework is particularly useful for investigating conserved synteny, identifying lineage-specific rearrangements, and exploring the evolutionary stability of gene clusters. The package outputs are designed to be analysis- and visualization-friendly, supporting downstream statistical analysis and graphical representation of gene neighborhood conservation.


# Installation
```{r, eval=FALSE}
BiocManager::install('jianhong/geneClusterPattern')
```

# Quick start
```{r}
library(geneClusterPattern)
library(org.Dr.eg.db)
library(GenomeInfoDb)
## prepare all the ensembl ids
ids <- as.list(org.Dr.egENSEMBL)
ensembl_gene_ids <- sort(unique(unlist(ids)))
```
```{r eval=FALSE}
## extract gene information via biomaRt
fish_mart <- guessSpecies('zebrafish', output='mart', version=112)
fish <- grangesFromEnsemblIDs(mart = fish_mart,
                              ensembl_gene_ids = ensembl_gene_ids)
```
```{r}
## subset the ensembl_gene_ids to save time
ensembl_gene_ids <- names(fish[seqnames(fish)=='24'])
## keep the standard sequence only
fish <- pruningSequences(fish)
## retrieve homologs
species <- guessSpecies(c('human', 'house mouse', 'Japanese medaka', 'turquoise killifish', 'gaculeatus'), version=112) # three-spined stickleback
species
```
```{r eval=FALSE}
homologs <- getHomologGeneList(species, fish_mart, ensembl_gene_ids)
homologs <- pruningSequences(homologs)
```
```{r}
## get gene cluster for target gene
queryGene <- 'pcolce2b'
nearest10neighbors <- getGeneCluster(fish, queryGene, homologs, k=10)

## plot cluster
genesList <- c(drerio=fish, homologs)[
  c("hsapiens", "mmusculus", "drerio", "olatipes", "nfurzeri", "gaculeatus")]

## plot
pgp <- plotGeneClusterPatterns(genesList, nearest10neighbors)
```


```{r}
gps <- geneOrderScore(genesList, ref='drerio', nearest10neighbors)
```


# From OrthoFinder results

You can start from the orthogroups table for one species of OrthorFinder.

```{r}
path <- system.file('extdata/orthofinder/Danio_rerio.GRCz11.pep.all.tsv.gz',
                    package='geneClusterPattern')
orthologs <- orthologPairsFromOrthoFinder(path)
annoGR_list <- readRDS(
    system.file('extdata/orthofinder/grange.obj.rds',
                 package='geneClusterPattern'))
homologs <- getHomologListForOrthoFinder(orthologs, annoGR_list[-1])
queryGR <- annoGR_list[["Danio rerio"]]
nearest10neighbors <- getGeneCluster(queryGR, queryGene, homologs, k=10)
pgp <- plotGeneClusterPatterns(c('zebrafish'=queryGR, homologs),
                               nearest10neighbors)
```

